Type Error Telemetry at Scale

### 2023-09-30 programming23/paper.tex
merge 4.9 with 4.8

\subsection{Do Edits Tend to Add Type Errors?}

@@

  \item
    In \FS{} mode, the percent of ``maintain'' records is high across
    the board and the percents of increases and decreases are lower.
    No matter the mode, reducing type errors is not guaranteed to reduce
    \FS{} errors as well.
    The next section further compares type and \FS{} errors.

@@

\subsection{Two Activities: Asset Creation and Scripting}
\label{s:strict-vs-forcedstrict}

\begin{figure}[t]\centering
  %% TODO
  %% - thin bars
  %% - colors for good , bad (disagree) , neutral
  \begin{tabular}{lll}
    \mnocheck{} & \mnonstrict{} & \mstrict{} \\
    \includegraphics[width=0.24\columnwidth]{img/compass-nocheck.pdf}
    &
    \includegraphics[width=0.24\columnwidth]{img/compass-nonstrict.pdf}
    &
    \includegraphics[width=0.24\columnwidth]{img/compass-strict.pdf}
  \end{tabular}
  \caption{Comparing the change in type errors to the change in \FS{} errors.}
  \label{f:tefs-compass}
\end{figure}

\Cref{f:tefs-compass} takes a closer look at type errors (\tekey{}) and \FS{} errors (\fskey{}).
\todo{I'm kinda' baffled.}
For every pair of events in each session, we compare the change in
type errors to the change in \FS{} errors and choose one of five categories.
At a high level, the question at hand is whether the changes \emph{agree} or \emph{disagree}.
The categories go one step further and separate zero and non-zero changes:

\noindent{}\begin{tabular}{lll}
  1. & \code{zero} &  neither \tekey{} nor \fskey{} changed \\
  2. & \code{te only} & \tekey{} changed but \fskey{} remained the same \\
  3. & \code{bg only} & \tekey{} remained the same but \fskey{} changed \\
  4. & \code{dis} & \tekey{} and \fskey{} changed in opposite directions \\
  5. & \code{agree} & \tekey{} and \fskey{} changed in the same direction
\end{tabular}
\smallskip{}

Of particular interest is how \mstrict{} and \FS{} checks compares.
The two analyses are similar, but \FS{} analysis uses strong checks
on all dependencies and has looser rules for the data model~(\cref{s:context}):
whereas \mstrict{} requires casts to confirm the types of data assets,
\FS{} gives data the dynamic gradual type.

\begin{itemize}
  \item
    All three plots have a nonzero percentage of \code{te only} items.
    In \mstrict{} mode, the percentage jumps up to nearly \pct{25}.
    Since \FS{} analysis is more lenient only for data model changes, these must
    be edits that affect the model in a way that raises a type error.
    The high percentage in \mstrict{} may be a source of frustration.
    %% Naturally, \mstrict{} mode has the largest number of such edits because
    %% its type checker is the most careful.

  \item
    In the first plot, for \mnocheck{},
    half the records are \code{zero}, nearly half are \code{bg only}, and only a few
    land in the other categories.
    While the number of zeros is a bit high, the high \code{bg only} and low
    \code{agree} counts make sense because \mnocheck{} reports few errors.

  \item
    In \mnonstrict{}, the number of agreements jumps up to nearly \pct{20}.
    This is reassuring because type and \FS{} analysis should agree to some extent,
    but the level of agreement is still rather low.

@@

This paper presents the first large-scale analysis of developers' interactions with
an industrial strength typechecker. It is nevertheless intentionally
limited by the nature and frequency of sampling. Despite these restrictions,
the analysis offers several lessons for the Luau
developers and other maintainers of gradually-typed languages:
\begin{itemize}
  \item
    Nearly all sessions (\pct{90}) use the default ``no check''
    typechecking mode~(\cref{f:dataset-overview}).
    Since the next typing mode (\mnonstrict{}) requires no annotations and is
    very conservative about the errors it reports, we conjecture that
    developers would have no objections to using it and are simply unaware that
    other modes exist.
%  \item
%    \pct{50} of all errors are in current module, so probably errors appear locally (not in a distant module);
%    \pct{5} of (current? old??) type errors overlap with edit range, so they don't stick around.
  \item
    Sessions tend to fix errors immediately rather than dealing with them in a batch
    or ignoring the errors altogether~(\cref{f:error-density,f:error-changes}).
    This should give Roblox confidence to make \mnonstrict{} the default in the future.
    % Changes to the number of errors balance one another~(\cref{f:error-density});
    % edits are equally likely to add or remove errors~(\cref{f:error-changes}).
  \item
    All Luau scripts interact with data assets, but no typechecking mode
    has a precise view of the data.
    Improving precision is an important direction for future work.
    Until then, \mstrict{} mode clearly needs to relax its data model
    analysis because it raises even more errors than \FS{} checking~(\cref{f:tefs-compass}).
\end{itemize}

On the meta level, the main takeaway is that the lightweight, counter-based telemetry
in Roblox Studio supports a variety of useful analyses:
\begin{itemize}
  \item
    Counting specific type errors helps to identify trends and to confirm that
    undesirable events rarely happen (e.g.,\code{UnificationTooComplex}).
  \item
    Testing the relationship between two counts can yield insights.
    The gap between \mstrict{} and \FS{} errors in~\cref{f:tefs-compass}
    revealed an issue with \mstrict{} checking.
  \item
    Reporting the overlap between type errors and the current (approximate) edit
    range lets us assess the outcome of edits without revealing source code.
    It was especially useful to have overlaps for two sets of type errors: old and current.
  \item
    Timestamps give useful, low-level insights about the frequency of keystrokes,
    the length of sessions, and actions over time.
    The event ordering that timestamps provide was invaluable; e.g., for learning
    that error deltas bounce up and down rather than rising shaply, then falling~(\cref{f:error-density}).
\end{itemize}

>>>>

\subsection{Do Edits Tend to Add Type Errors?}
\label{s:tefs-compass}

@@

  \item
    In \FS{} mode, the percent of ``maintain'' records is high across
    the board and the increases and decreases are lower.
    No matter the mode, reducing type errors is not guaranteed to reduce
    \FS{} errors as well.
    This is surprising for \mstrict{} mode, and likely relates to
    the data model; see below for details.

@@

\paragraph{Data Model Types: Strict vs. Background Analysis}

Based on the description of \FS{} analysis from~\cref{s:context},
it should raise \emph{at least as many errors} as \mstrict{} mode
because it uses \mstrict{} checks for all dependencies.
For example, if a \mstrict{} module imports from a \mnocheck{} one,
then \mstrict{} analysis does not analyze the import but
\FS{} analysis does.

In \cref{f:error-changes}, however, nearly \pct{3} of \mstrict{}-mode records
increase the number of \mstrict{} errors without increasing \FS{} errors.
Further inspection reveals that the situation is a bit worse, as
some records increase \mstrict{} errors without changing the \FS{} errors:
\begin{itemize}
  \item
    \pct{22} of \mstrict{} records increase (\addsym{}) the number of type errors
    but maintain (\keepsym{}) the number of \FS{} errors.
\end{itemize}
The reason for the discrepancy is that \mstrict{} mode and \FS{} analysis
assign different types to the data model: \mstrict{} assigns a top type
and requires downcasts at use sites, whereas \FS{} analysis uses the
dynamic gradual type.
Dealing with these data model errors may be a source of frustration
for creators using \mstrict{} mode.

@@

This paper presents the first large-scale analysis of developers' interactions with
an industrial strength typechecker. It is nevertheless intentionally
limited by the nature and frequency of sampling. Despite these restrictions,
the analysis offers several lessons for the Luau
developers and other maintainers of gradually-typed languages:
\begin{itemize}
  \item
    Nearly all sessions (\pct{90}) use the default ``no check''
    typechecking mode~(\cref{f:dataset-overview}).
    Since the next typing mode (\mnonstrict{}) requires no annotations and is
    very conservative about the errors it reports, we conjecture that
    developers would have no objections to using it and are simply unaware that
    other modes exist.
%  \item
%    \pct{50} of all errors are in current module, so probably errors appear locally (not in a distant module);
%    \pct{5} of (current? old??) type errors overlap with edit range, so they don't stick around.
  \item
    Sessions tend to fix errors immediately rather than dealing with them in a batch
    or ignoring the errors altogether~(\cref{f:error-density,f:error-changes}).
    This should give Roblox confidence to make \mnonstrict{} the default in the future.
    % Changes to the number of errors balance one another~(\cref{f:error-density});
    % edits are equally likely to add or remove errors~(\cref{f:error-changes}).
  \item
    All Luau scripts interact with data assets, but no typechecking mode
    has a precise view of the data.
    Improving precision is an important direction for future work.
    Until then, \mstrict{} mode clearly needs to relax its data model
    analysis because it raises even more errors than \FS{} checking~(\cref{s:tefs-compass}).
\end{itemize}

On the meta level, the main takeaway is that the lightweight, counter-based telemetry
in Roblox Studio supports a variety of useful analyses:
\begin{itemize}
  \item
    Counting specific type errors helps to identify trends and to confirm that
    undesirable events rarely happen (e.g.,\code{UnificationTooComplex}).
  \item
    Testing the relationship between two counts can yield insights.
    The gap between \mstrict{} and \FS{} errors in~\cref{s:tefs-compass}
    revealed an issue with \mstrict{} checking.
  \item
    Reporting the overlap between type errors and the current (approximate) edit
    range lets us assess the outcome of edits without revealing source code.
    It was especially useful to have overlaps for two sets of type errors: old and current.
  \item
    Timestamps give useful, low-level insights about the frequency of keystrokes,
    the length of sessions, and actions over time.
    The event ordering that timestamps provide was invaluable; e.g., for learning
    that error deltas bounce up and down rather than rising shaply, then falling~(\cref{f:error-density}).
\end{itemize}
